### 2019-04-24 Chapters/Chapter3.tex
cdots and spaces

\begin{equation}
\label{Ham1}
\hat{H} = \sum_{i=1}^{N_e} \hat{H}^{(1)}(\textbf{r}_i) + \frac{1}{2} \sum_{\substack{i,j = 1 \\ i \neq j}} ^ {N_e} v(\textbf{r}_i - \textbf{r}_j)
\end{equation}

where $v$ is the electron-electron interaction and where $\hat{H}^{(1)}$ is the single particle Hamiltonian:

\begin{equation}
\hat{H}^{(1)}(\textbf{r}) = -\frac{\hbar^2}{2m}\nabla^2_{\textbf{r}} + \text{U}(\textbf{r}) + \lambda\bs{\sigma}\cdot\bs{\nabla}V(\textbf{r}) \times \textbf{p}
\end{equation}

Where $\lambda = \frac{\hbar^2}{4m^2c^2}$ is the spin orbit coupling constant and $\text{U}(\bs{r})$ is a periodic potential. The second quantized form of this Hamiltonian can be approximated by a version of the Hubbard model. We will consider a half filled system in the strong coupling limit. Ignoring electron hopping and spin orbit interaction, the ground state of this system is massively degenerated. We will find an effective Hamiltonian considering that the hopping term and the spin interaction term act on this subspace. This effective Hamiltonian will be the sum of three spin interactions: an exchange interaction $J\bs{S}_i\cdot\bs{S}_j$ with $J>0$, a Dzyaloshinkii-Moriya interaction $\bs{D}\cdot\bs{S}_i\times\bs{S}_j$ and an anisotropic interaction $\bs{S}_i \bs{\Gamma} \bs{S}_j$ known as pseudodipolar interaction \cite{Moriya1960}. We will start by finding a suitable basis to write the second quantized version of \ref{Ham1}.

@@

\begin{equation}
\label{Bloch1}
\bra{\bs{r}+\bs{R}} \ket{\phi_{n, \textbf{k}}} = e^{i\bs{k}\cdot\bs{R}}\bra{\bs{r}} \ket{\phi_{n, \textbf{k}}}
\end{equation}

Which is the well-known Bloch function form. Here, $n$ is called the band index, and $\textbf{k}$ is the quasimomentum. The energy of this wavefunction is $\epsilon_{n \textbf{k}}$.  Notice that both $\phi_{n,\textbf{k}}$ and $\epsilon_{n \textbf{k}}$ are periodic functions of $\textbf{k}$ in the reciprocal lattice.

@@

\begin{align}
\bra{\bs{r}}\ket{\phi_{n\bs{k}}} &= \frac{1}{\sqrt{M}}\sum_{\bs{R}} e^{i\bs{k}\cdot\bs{R}} \bra{\bs{r}}\ket{\psi_{\bs{R}n}}  \label{Wannier1} \\
\bra{\bs{r}}\ket{\psi_{in}} &= \frac{1}{\sqrt{M}} \sum_{\bs{k}\in BZ} e^{-i\bs{k}\cdot\bs{R}_i} \bra{\bs{r}}\ket{\phi_{n\bs{k}}}\label{Wannier2}
\end{align}

Where $M$ is the number of lattice sites as defined in \ref{AP1A}. Using \ref{Bloch1} we see that $\psi_{in}(\textbf{r}) = \psi_{0n}(\textbf{r}-\textbf{r}_i) \equiv \psi_{n}(\textbf{r}-\textbf{r}_i)$, so we only need to define one Wannier function for each band and the others are obtained by translations. From now on we will restrict ourselves to a fixed band, so we will drop the band index. Additionally, we will include the spin state, therefore the basis states will have two quantum numbers one for the site and one for the spin state. We will adopt the notation $\ket{i\sigma}$ for the basis state of site $i$ with spin state $\sigma = \uparrow, \downarrow$, or just $\ket{i}$ when the spin state is irrelevant. Fock states with small number of particles will be denoted as $\ket{i\sigma, j\sigma'}$, for a state with a particle at site $i$ with spin $\sigma$ and a particle at $j$ with spin $\sigma'$, or $\ket{ij}$ when spin is irrelevant. With these definitions we are ready for write down the second quantized form of \ref{Ham1}, and show some approximations that lead to the Hubbard model.

@@

\begin{equation}
\hat{H} = \sum_{i,j,\sigma, \sigma'} \bra{i\sigma} \hat{H}^{(1)} \ket{j\sigma'} \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'} + \frac{1}{2} \sum_{i,j,k,l, \sigma, \sigma'} \bra{ij} \hat{v} \ket{kl} \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'}^\dagger \hat{c}_{l \sigma'} \hat{c}_{k \sigma}
\end{equation}

Where we assumed $\hat{v}$ to be spin independent. This can be rewritten as:

\begin{equation}
\hat{H} = -\sum_{i,j,\sigma, \sigma'} (\delta_{\sigma, \sigma'} t_{ij} + \bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma, \sigma'}) \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'} + \frac{1}{2} \sum_{i,j,k,l, \sigma, \sigma'} \bra{ij} \hat{v} \ket{kl} \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'}^\dagger \hat{c}_{l \sigma'} \hat{c}_{k \sigma}
\end{equation}

Where $t_{ij} = -\bra{i} -\frac{\hbar^2}{2m}\nabla^2_{\textbf{r}} + \text{U}(\textbf{r}) \ket{j}$ and $\boldsymbol{\Delta}_{ij} = -\bra{i} \lambda \boldsymbol{\nabla}V(\textbf{r}) \times \textbf{p} \ket{j}$. Notice that $\bs{\Delta}_{ij} = - \bs{\Delta}_{ji}$, for example, in the $x$ component:

\begin{align*}
\Delta_{ij,x} &= -\bra{i} \lambda (\partial_y V(\bs{r}) p_z - \partial_z V(\bs{r}) p_y) \ket{j} = \\
&= i \hbar \lambda \int d \bs{r} \phi_i(\bs{r})^* (\partial_y V(\bs{r}) \partial_z - \partial_z V(\bs{r}) \partial_y) \phi_j(\bs{r}) = \\
&= -i\hbar\lambda \int d\bs{r} \left[ \partial_z(\phi_i(\bs{r})^*\partial_y V(\bs{r}) - \partial_y(\phi_i(\bs{r})^*\partial_z V(\bs{r}) \right] \phi_j(\bs{r}) = \\
&= -i\hbar\lambda \int d\bs{r} \left[ \partial_z \phi_i(\bs{r})^* \partial_y V(\bs{r}) - \partial_y \phi_i(\bs{r})^*\partial_z V(\bs{r}) \right] \phi_j(\bs{r}) = \\
&= \int d\bs{r} \phi_j(\bs{r}) \lambda (\bs{\nabla}V(\bs{r}) \times \bs{p})_x \phi_i(\bs{r})^* = - \Delta_{ji,x}
\end{align*}

Therefore it follows that $\bs{\Delta}_{ii} = 0$ and that $\bs{\Delta}_{ij}$ is purely imaginary. We can impose that $t_{ii} = t(0) = 0$ since adding a constant term to all the hopping amplitudes would only lead to a lifting in the zero point energy. Additionally, for a system of strongly correlated electrons we can assume that:

@@

\begin{equation}
\bs{\Delta}_{ij} = \begin{cases}
             \bs{\Delta}_{ij},  & \text{for } (i,j) \text{ nearest neighbours} \\
             0,  & \text{otherwise}
       \end{cases} \quad
\end{equation}

Hubbard models with hopping terms further than nearest neighbors are also interesting and will be considered in next chapter.

@@

\begin{equation}
\bra{ij} \hat{v} \ket{kl} =
	\begin{cases}
		\text{U}, & \text{if } i=j=k=l \\
		\text{V}, & \text{if } i=k \text{ and } j=l \text{ where } (i,j) \text{ nearest neighbours} \\
		0, & \text{otherwise}
	\end{cases}						
\end{equation}

Where we defined:

\begin{align}
\text{U} &= \frac{1}{2} \int d\bs{r} d\bs{r}' \phi(\bs{r})^* \phi(\bs{r}')^* \hat{v}(\bs{r}-\bs{r}') \phi(\bs{r}') \phi(\bs{r}) \\
\text{V} &= \int d\bs{r} d\bs{r}' \phi(\bs{r})^* \phi(\bs{r}'-\bs{R}_1)^* \hat{v}(\bs{r}-\bs{r}') \phi(\bs{r}'-\bs{R}_1) \phi(\bs{r})
\end{align}

Where $\phi(\bs{r})$ is the Wannier function centered at the origin and $\bs{R}_1$ is a nearest neighbour lattice vector. Notice that we made explicit use of the translational invariance of the system. Also, for $i=j=k=l$ the Pauli principle requires $\sigma' = -\sigma$.

@@

\begin{equation}
\label{HubbardWithNonLocalInt}
\hat{H} = -\sum_{\langle i,j \rangle, \sigma, \sigma'}(\delta_{\sigma, \sigma'} t_{ij} + \bs{\Delta}_{ij} \cdot \bs{\sigma}_{\sigma, \sigma'})\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'} + \text{U} \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow} + \text{V} \sum_{i,j,\sigma,\sigma'}  \hat{n}_{i \sigma} \hat{n}_{j \sigma'}
\end{equation}

Now, in a half filling system this Hamiltonian can be well approximated by an effective Hamiltonian without non local Coulomb interactions:

\begin{equation}
\label{Intermediate}
\hat{H} = -\sum_{\langle i,j \rangle, \sigma, \sigma'}(\delta_{\sigma, \sigma'} t_{ij} + \bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma, \sigma'})\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'} + \text{U}^* \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}
\end{equation}

Where $\text{U}^* = \text{U} - \text{V}$. The physical reason for this is that the energy difference when the Coulomb interaction pushes an electron from a double occupied site to an empty site is $\text{U}-\text{V}$ in \ref{HubbardWithNonLocalInt}, because the number of double occupied sites is reduced by one but the number of adjacent occupied sites is increased by one. This is illustred in Figure \ref{Fig2.1}. A more rigouros derivation of this effective model can be found in \cite{Schuler2013}. Altough $\text{V} < \text{U}$, they can be comparable in magnitude, and $\text{U}^*$ can be reduced by a factor of $0.5$ or more.

@@

\begin{equation}
\label{Hubbard}
\hat{H} = -\sum_{\langle i,j \rangle, \sigma, \sigma'}(\delta_{\sigma, \sigma'} t_{ij} + \bs{\Delta}_{ij} \cdot \bs{\sigma}_{\sigma, \sigma'})\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'} + \text{U} \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}
\end{equation}

This Hamiltonian is identical to the Hubbard Hamiltonian \ref{HH} adding a spin dependent hopping, which arises from the SOI. In the next section we will show that this Hamiltonian leads to an effective spin Hamiltonian in the appropriate limit.

@@

\begin{equation}
\bra{\phi'} \hat{H} \ket{\phi} = \sum_s \frac{\bra{\phi'} \hat{H_t} \ket{s}\bra{s} \hat{H_t} \ket{\phi}}{E_0-E_s}
\end{equation}

The sum being over all excited states. Since $\hat{H_t}\ket{\phi}$ is a superposition of states with exactly one double occupied site and one empty site, we see that only excited states with one double occupied site will contribute. These states have energy $E_s = \text{U}$ therefore we have $\bra{\phi'} \hat{H} \ket{\phi} = -\frac{1}{\text{U}}\sum_s \bra{\phi'} \hat{H_t} \ket{s}\bra{s} \hat{H_t} \ket{\phi} = \bra{\phi'} -\frac{\hat{H_t^2}}{\text{U}} \ket{\phi}$. This means that within the subspace generated by the ground states of $\hat{H_0}$, $\hat{H}$ acts as an effective Hamiltonian $\hat{H}_{\text{eff}} = -\frac{\hat{H_t^2}}{\text{U}}$. Now,

\begin{align}
\label{PertNoEMF}
&\bra{\phi'} -\frac{\hat{H_t^2}}{\text{U}} \ket{\phi} = \nonumber \\
&= -\frac{1}{\text{U}} \bra{\phi'}\sum_{\langle i,j \rangle, \sigma_1, \sigma_2} \sum_{\langle i',j' \rangle, 					\sigma_3, \sigma_4} \left\{ \left( \delta_{\sigma_1, \sigma_2} t + \bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma_1, \sigma_2} \right) \left( \delta_{\sigma_3, \sigma_4} t + \bs{\Delta}_{i'j'}\cdot \bs{\sigma}_{\sigma_3, \sigma_4} \right) \right. \nonumber \\
& \left. \hat{c}_{i\sigma_1}^\dagger \hat{c}_{j \sigma_2} \hat{c}_{i' \sigma_3}^\dagger \hat{c}_{j'\sigma_4} \ket{\phi} \right\} = \nonumber \\
&= -\frac{1}{\text{U}} \bra{\phi'}\sum_{\langle i,j \rangle, \sigma_1, \sigma_2, \sigma_3, \sigma_4} \left( \delta_{\sigma_1, \sigma_2} t + \bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma_1, \sigma_2} \right) \left( \delta_{\sigma_3, \sigma_4} t - \bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma_3, \sigma_4} \right)\hat{c}_{i\sigma_1}^\dagger \hat{c}_{j \sigma_2} \hat{c}_{j \sigma_3}^\dagger \hat{c}_{i\sigma_4} \ket{\phi} = \nonumber \\
&-\frac{1}{\text{U}} \bra{\phi'} \sum_{\langle i,j \rangle, \sigma_1, \sigma_2, \sigma_3, \sigma_4} \left\{ \left[ t^2 						\delta_{\sigma_1 \sigma_2}\delta_{\sigma_3 \sigma_4} + t\bs{\Delta}_{ij}(\delta_{\sigma_3 \sigma_4}			\bs{\sigma}_{\sigma_1 \sigma_2} \right. \right. \nonumber \\
& \left. \left. - \delta_{\sigma_1 \sigma_2} \bs{\sigma}_{\sigma_3 \sigma_4}) - (\bs{\Delta}_{ij}\cdot\bs{\sigma}_{\sigma_1 \sigma_2})(\bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma_3 \sigma_4}) \right] \hat{c}_{i\sigma_1}^\dagger \hat{c}_{i\sigma_4} \hat{c}_{j 		\sigma_2} \hat{c}_{j \sigma_3}^\dagger \ket{\phi} \right\} = \nonumber \\
& \bra{\phi'} \hat{H}_J + \hat{H}_D + \hat{H}_P \ket{\phi}
\end{align}

In the second line we imposed $i=j'$ and $j=i'$ since that's the only way $\hat{H_t^2}\ket{\phi}$ has no double occupied sites and so the matrix element does not vanish (remember that $\ket{\phi}$ and $\ket{\phi'}$ belong in the subspace of zero double occupied states). We also used $\bs{\Delta}_{ij} = - \bs{\Delta}_{ji}$ and we defined:

\begin{align}
\hat{H}_J &= -\frac{t^2}{\text{U}} \sum_{\langle i,j \rangle, \sigma, \sigma'} \hat{c}_{i\sigma}^\dagger \hat{c}_{i\sigma'} \hat{c}_{j\sigma} \hat{c}_{j \sigma'}^\dagger \label{C3HJ} \\
\hat{H}_D &= -\frac{t}{\text{U}}\sum_{\langle i,j \rangle, \sigma_1, \sigma_2, \sigma_3, \sigma_4} \bs{\Delta}_{ij}(\delta_{\sigma_3 \sigma_4}	\bs{\sigma}_{\sigma_1 \sigma_2} - \delta_{\sigma_1 \sigma_2} \bs{\sigma}_{\sigma_3 	\sigma_4}) \hat{c}_{i\sigma_1}^\dagger \hat{c}_{i\sigma_4} \hat{c}_{j \sigma_2} \hat{c}_{j \sigma_3}^\dagger \label{C3HD}\\
\hat{H}_P &= \frac{1}{\text{U}}\sum_{\langle i,j \rangle, \sigma_1, \sigma_2, \sigma_3, \sigma_4} (\bs{\Delta}_{ij}\bs{\sigma}_{\sigma_1 \sigma_2})(\bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma_3 \sigma_4}) \hat{c}_{i\sigma_1}^\dagger \hat{c}_{i\sigma_4} \hat{c}_{j \sigma_2} \hat{c}_{j \sigma_3}^\dagger \label{C3HP}
\end{align}

Let us now introduce the on-site spin operators:

\begin{equation}
\label{SpinOperators}
\boldsymbol{S}_i = \frac{1}{2} \sum_{\sigma, \sigma'} \hat{c}_{i \sigma}^\dagger \boldsymbol{\sigma}_{\sigma, \sigma'} \hat{c}_{i \sigma'}
\end{equation}

Which satisfy:

>>>>

\begin{equation}
\label{Ham1}
\hat{H} = \sum_{i=1}^{N_e} \hat{H}^{(1)}(\textbf{r}_i) + \frac{1}{2} \sum_{\substack{i,j = 1 \\ i \neq j}} ^ {N_e} v(\textbf{r}_i - \textbf{r}_j)
\end{equation}
where $v$ is the electron-electron interaction and where $\hat{H}^{(1)}$ is the single particle Hamiltonian:

\begin{equation}
\hat{H}^{(1)}(\textbf{r}) = -\frac{\hbar^2}{2m}\nabla^2_{\textbf{r}} + \text{U}(\textbf{r}) + \lambda\bs{\sigma}\cdot\bs{\nabla}V(\textbf{r}) \times \textbf{p}
\end{equation}
Where $\lambda = \frac{\hbar^2}{4m^2c^2}$ is the spin orbit coupling constant and $\text{U}(\bs{r})$ is a periodic potential. $\hbar$ is the reduced Planck constant, $m$ is the electron mass, $c$ is the speed of light and $\bs{\sigma}$ is a vector of Pauli matrices $\bs{\sigma} = (\sigma_1, \sigma_2, \sigma_3)$. The second quantized form of this Hamiltonian can be approximated by a version of the Hubbard model. We will consider a half filled system in the strong coupling limit. Ignoring electron hopping and spin orbit interaction, the ground state of this system is massively degenerated. We will find an effective Hamiltonian considering that the hopping term and the spin interaction term act on this subspace. This effective Hamiltonian will be the sum of three spin interactions: an exchange interaction $J\bs{S}_i\cdot\bs{S}_j$ with $J>0$, a Dzyaloshinkii-Moriya interaction $\bs{D}\cdot\bs{S}_i\times\bs{S}_j$ and an anisotropic interaction $\bs{S}_i \bs{\Gamma} \bs{S}_j$ known as pseudodipolar interaction \cite{Moriya1960}. We will start by finding a suitable basis to write the second quantized version of \ref{Ham1}.

@@

\begin{equation}
\label{Bloch1}
\bra{\bs{r}+\bs{R}} \ket{\phi_{n, \textbf{k}}} = e^{i\bs{k}\cdot\bs{R}}\bra{\bs{r}} \ket{\phi_{n, \textbf{k}}}
\end{equation}
Which is the well-known Bloch function form. Here, $n$ is called the band index, and $\textbf{k}$ is the quasimomentum. The energy of this wavefunction is $\epsilon_{n \textbf{k}}$.  Notice that both $\phi_{n,\textbf{k}}$ and $\epsilon_{n \textbf{k}}$ are periodic functions of $\textbf{k}$ in the reciprocal lattice.

@@

\begin{align}
\bra{\bs{r}}\ket{\phi_{n\bs{k}}} &= \frac{1}{\sqrt{M}}\sum_{\bs{R}} e^{i\bs{k}\cdot\bs{R}} \bra{\bs{r}}\ket{\psi_{\bs{R}n}}  \label{Wannier1} \\
\bra{\bs{r}}\ket{\psi_{in}} &= \frac{1}{\sqrt{M}} \sum_{\bs{k}\in BZ} e^{-i\bs{k}\cdot\bs{R}_i} \bra{\bs{r}}\ket{\phi_{n\bs{k}}}\label{Wannier2}
\end{align}
Where $M$ is the number of lattice sites as defined in \ref{AP1A}. Using \ref{Bloch1} we see that $\psi_{in}(\textbf{r}) = \psi_{0n}(\textbf{r}-\textbf{r}_i) \equiv \psi_{n}(\textbf{r}-\textbf{r}_i)$, so we only need to define one Wannier function for each band and the others are obtained by translations. From now on we will restrict ourselves to a fixed band, so we will drop the band index. Additionally, we will include the spin state, therefore the basis states will have two quantum numbers one for the site and one for the spin state. We will adopt the notation $\ket{i\sigma}$ for the basis state of site $i$ with spin state $\sigma = \uparrow, \downarrow$, or just $\ket{i}$ when the spin state is irrelevant. Fock states with small number of particles will be denoted as $\ket{i\sigma, j\sigma'}$, for a state with a particle at site $i$ with spin $\sigma$ and a particle at $j$ with spin $\sigma'$, or $\ket{ij}$ when spin is irrelevant. With these definitions we are ready for write down the second quantized form of \ref{Ham1}, and show some approximations that lead to the Hubbard model.

@@

\begin{equation}
\hat{H} = \sum_{i,j,\sigma, \sigma'} \bra{i\sigma} \hat{H}^{(1)} \ket{j\sigma'} \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'} + \frac{1}{2} \sum_{i,j,k,l, \sigma, \sigma'} \bra{ij} \hat{v} \ket{kl} \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'}^\dagger \hat{c}_{l \sigma'} \hat{c}_{k \sigma}
\end{equation}
Where we assumed $\hat{v}$ to be spin independent. This can be rewritten as:

\begin{equation}
\hat{H} = -\sum_{i,j,\sigma, \sigma'} (\delta_{\sigma, \sigma'} t_{ij} + \bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma, \sigma'}) \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'} + \frac{1}{2} \sum_{i,j,k,l, \sigma, \sigma'} \bra{ij} \hat{v} \ket{kl} \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'}^\dagger \hat{c}_{l \sigma'} \hat{c}_{k \sigma}
\end{equation}
Where $t_{ij} = -\bra{i} -\frac{\hbar^2}{2m}\nabla^2_{\textbf{r}} + \text{U}(\textbf{r}) \ket{j}$ and $\boldsymbol{\Delta}_{ij} = -\bra{i} \lambda \boldsymbol{\nabla}V(\textbf{r}) \times \textbf{p} \ket{j}$. Notice that $\bs{\Delta}_{ij} = - \bs{\Delta}_{ji}$, for example, in the $x$ component:

\begin{align*}
\Delta_{ij,x} &= -\bra{i} \lambda (\partial_y V(\bs{r}) p_z - \partial_z V(\bs{r}) p_y) \ket{j} = \\
&= i \hbar \lambda \int d \bs{r} \phi_i(\bs{r})^* (\partial_y V(\bs{r}) \partial_z - \partial_z V(\bs{r}) \partial_y) \phi_j(\bs{r}) = \\
&= -i\hbar\lambda \int d\bs{r} \left[ \partial_z(\phi_i(\bs{r})^*\partial_y V(\bs{r}) - \partial_y(\phi_i(\bs{r})^*\partial_z V(\bs{r}) \right] \phi_j(\bs{r}) = \\
&= -i\hbar\lambda \int d\bs{r} \left[ \partial_z \phi_i(\bs{r})^* \partial_y V(\bs{r}) - \partial_y \phi_i(\bs{r})^*\partial_z V(\bs{r}) \right] \phi_j(\bs{r}) = \\
&= \int d\bs{r} \phi_j(\bs{r}) \lambda (\bs{\nabla}V(\bs{r}) \times \bs{p})_x \phi_i(\bs{r})^* = - \Delta_{ji,x}
\end{align*}
Therefore it follows that $\bs{\Delta}_{ii} = 0$ and that $\bs{\Delta}_{ij}$ is purely imaginary. We can impose that $t_{ii} = t(0) = 0$ since adding a constant term to all the hopping amplitudes would only lead to a lifting in the zero point energy. Additionally, for a system of strongly correlated electrons we can assume that:

@@

\begin{equation}
\bs{\Delta}_{ij} = \begin{cases}
             \bs{\Delta}_{ij},  & \text{for } (i,j) \text{ nearest neighbours} \\
             0,  & \text{otherwise}
       \end{cases} \quad
\end{equation}
Hubbard models with hopping terms further than nearest neighbors are also interesting and will be considered in next chapter.

@@

\begin{equation}
\bra{ij} \hat{v} \ket{kl} =
	\begin{cases}
		\text{U}, & \text{if } i=j=k=l \\
		\text{V}, & \text{if } i=k \text{ and } j=l \text{ where } (i,j) \text{ nearest neighbours} \\
		0, & \text{otherwise}
	\end{cases}						
\end{equation}
Where we defined:

\begin{align}
\text{U} &= \frac{1}{2} \int d\bs{r} d\bs{r}' \phi(\bs{r})^* \phi(\bs{r}')^* \hat{v}(\bs{r}-\bs{r}') \phi(\bs{r}') \phi(\bs{r}) \\
\text{V} &= \int d\bs{r} d\bs{r}' \phi(\bs{r})^* \phi(\bs{r}'-\bs{R}_1)^* \hat{v}(\bs{r}-\bs{r}') \phi(\bs{r}'-\bs{R}_1) \phi(\bs{r})
\end{align}
Where $\phi(\bs{r})$ is the Wannier function centered at the origin and $\bs{R}_1$ is a nearest neighbour lattice vector. Notice that we made explicit use of the translational invariance of the system. Also, for $i=j=k=l$ the Pauli principle requires $\sigma' = -\sigma$.

@@

\begin{equation}
\label{HubbardWithNonLocalInt}
\hat{H} = -\sum_{\langle i,j \rangle, \sigma, \sigma'}(\delta_{\sigma, \sigma'} t_{ij} + \bs{\Delta}_{ij} \cdot \bs{\sigma}_{\sigma, \sigma'})\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'} + \text{U} \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow} + \text{V} \sum_{i,j,\sigma,\sigma'}  \hat{n}_{i \sigma} \hat{n}_{j \sigma'}
\end{equation}
Now, in a half filling system this Hamiltonian can be well approximated by an effective Hamiltonian without non local Coulomb interactions:

\begin{equation}
\label{Intermediate}
\hat{H} = -\sum_{\langle i,j \rangle, \sigma, \sigma'}(\delta_{\sigma, \sigma'} t_{ij} + \bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma, \sigma'})\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'} + \text{U}^* \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}
\end{equation}
Where $\text{U}^* = \text{U} - \text{V}$. The physical reason for this is that the energy difference when the Coulomb interaction pushes an electron from a double occupied site to an empty site is $\text{U}-\text{V}$ in \ref{HubbardWithNonLocalInt}, because the number of double occupied sites is reduced by one but the number of adjacent occupied sites is increased by one. This is illustred in Figure \ref{Fig2.1}. A more rigouros derivation of this effective model can be found in \cite{Schuler2013}. Altough $\text{V} < \text{U}$, they can be comparable in magnitude, and $\text{U}^*$ can be reduced by a factor of $0.5$ or more.

@@

\begin{equation}
\label{Hubbard}
\hat{H} = -\sum_{\langle i,j \rangle, \sigma, \sigma'}(\delta_{\sigma, \sigma'} t_{ij} + \bs{\Delta}_{ij} \cdot \bs{\sigma}_{\sigma, \sigma'})\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'} + \text{U} \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}
\end{equation}
This Hamiltonian is identical to the Hubbard Hamiltonian \ref{HH} adding a spin dependent hopping, which arises from the SOI. In the next section we will show that this Hamiltonian leads to an effective spin Hamiltonian in the appropriate limit.

@@

\begin{equation}
\bra{\phi'} \hat{H} \ket{\phi} = \sum_s \frac{\bra{\phi'} \hat{H_t} \ket{s}\bra{s} \hat{H_t} \ket{\phi}}{E_0-E_s}
\end{equation}
The sum being over all excited states. Since $\hat{H_t}\ket{\phi}$ is a superposition of states with exactly one double occupied site and one empty site, we see that only excited states with one double occupied site will contribute. These states have energy $E_s = \text{U}$ therefore we have $\bra{\phi'} \hat{H} \ket{\phi} = -\frac{1}{\text{U}}\sum_s \bra{\phi'} \hat{H_t} \ket{s}\bra{s} \hat{H_t} \ket{\phi} = \bra{\phi'} -\frac{\hat{H_t^2}}{\text{U}} \ket{\phi}$. This means that within the subspace generated by the ground states of $\hat{H_0}$, $\hat{H}$ acts as an effective Hamiltonian $\hat{H}_{\text{eff}} = -\frac{\hat{H_t^2}}{\text{U}}$. Now,

\begin{align}
\label{PertNoEMF}
&\bra{\phi'} -\frac{\hat{H_t^2}}{\text{U}} \ket{\phi} = \nonumber \\
&= -\frac{1}{\text{U}} \bra{\phi'}\sum_{\langle i,j \rangle, \sigma_1, \sigma_2} \sum_{\langle i',j' \rangle, 					\sigma_3, \sigma_4} \left\{ \left( \delta_{\sigma_1, \sigma_2} t + \bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma_1, \sigma_2} \right) \left( \delta_{\sigma_3, \sigma_4} t + \bs{\Delta}_{i'j'}\cdot \bs{\sigma}_{\sigma_3, \sigma_4} \right) \right. \nonumber \\
& \left. \hat{c}_{i\sigma_1}^\dagger \hat{c}_{j \sigma_2} \hat{c}_{i' \sigma_3}^\dagger \hat{c}_{j'\sigma_4} \ket{\phi} \right\} = \nonumber \\
&= -\frac{1}{\text{U}} \bra{\phi'}\sum_{\langle i,j \rangle, \sigma_1, \sigma_2, \sigma_3, \sigma_4} \left( \delta_{\sigma_1, \sigma_2} t + \bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma_1, \sigma_2} \right) \left( \delta_{\sigma_3, \sigma_4} t - \bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma_3, \sigma_4} \right)\hat{c}_{i\sigma_1}^\dagger \hat{c}_{j \sigma_2} \hat{c}_{j \sigma_3}^\dagger \hat{c}_{i\sigma_4} \ket{\phi} = \nonumber \\
&-\frac{1}{\text{U}} \bra{\phi'} \sum_{\langle i,j \rangle, \sigma_1, \sigma_2, \sigma_3, \sigma_4} \left\{ \left[ t^2 						\delta_{\sigma_1 \sigma_2}\delta_{\sigma_3 \sigma_4} + t\bs{\Delta}_{ij}(\delta_{\sigma_3 \sigma_4}			\bs{\sigma}_{\sigma_1 \sigma_2} \right. \right. \nonumber \\
& \left. \left. - \delta_{\sigma_1 \sigma_2} \bs{\sigma}_{\sigma_3 \sigma_4}) - (\bs{\Delta}_{ij}\cdot\bs{\sigma}_{\sigma_1 \sigma_2})(\bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma_3 \sigma_4}) \right] \hat{c}_{i\sigma_1}^\dagger \hat{c}_{i\sigma_4} \hat{c}_{j 		\sigma_2} \hat{c}_{j \sigma_3}^\dagger \ket{\phi} \right\} = \nonumber \\
& \bra{\phi'} \hat{H}_J + \hat{H}_D + \hat{H}_P \ket{\phi}
\end{align}
In the second line we imposed $i=j'$ and $j=i'$ since that's the only way $\hat{H_t^2}\ket{\phi}$ has no double occupied sites and so the matrix element does not vanish (remember that $\ket{\phi}$ and $\ket{\phi'}$ belong in the subspace of zero double occupied states). We also used $\bs{\Delta}_{ij} = - \bs{\Delta}_{ji}$ and we defined:

\begin{align}
\hat{H}_J &= -\frac{t^2}{\text{U}} \sum_{\langle i,j \rangle, \sigma, \sigma'} \hat{c}_{i\sigma}^\dagger \hat{c}_{i\sigma'} \hat{c}_{j\sigma} \hat{c}_{j \sigma'}^\dagger \label{C3HJ} \\
\hat{H}_D &= -\frac{t}{\text{U}}\sum_{\langle i,j \rangle, \sigma_1, \sigma_2, \sigma_3, \sigma_4} \bs{\Delta}_{ij}(\delta_{\sigma_3 \sigma_4}	\bs{\sigma}_{\sigma_1 \sigma_2} - \delta_{\sigma_1 \sigma_2} \bs{\sigma}_{\sigma_3 	\sigma_4}) \hat{c}_{i\sigma_1}^\dagger \hat{c}_{i\sigma_4} \hat{c}_{j \sigma_2} \hat{c}_{j \sigma_3}^\dagger \label{C3HD}\\
\hat{H}_P &= \frac{1}{\text{U}}\sum_{\langle i,j \rangle, \sigma_1, \sigma_2, \sigma_3, \sigma_4} (\bs{\Delta}_{ij}\bs{\sigma}_{\sigma_1 \sigma_2})(\bs{\Delta}_{ij} \cdot\bs{\sigma}_{\sigma_3 \sigma_4}) \hat{c}_{i\sigma_1}^\dagger \hat{c}_{i\sigma_4} \hat{c}_{j \sigma_2} \hat{c}_{j \sigma_3}^\dagger \label{C3HP}
\end{align}
Let us now introduce the on-site spin operators:

\begin{equation}
\label{SpinOperators}
\boldsymbol{S}_i = \frac{1}{2} \sum_{\sigma, \sigma'} \hat{c}_{i \sigma}^\dagger \boldsymbol{\sigma}_{\sigma, \sigma'} \hat{c}_{i \sigma'}
\end{equation}
Which satisfy:
